\documentclass{slides}
\usepackage{tikz-uml}

\lstset{language=C++}

\begin{document}

\graphicspath{{figures/}}

\title[Object-oriented Programming in C++]{\Large Object-oriented
  Programming in C++}

\author[A. Arnold and O. Lenz]{Axel Arnold \and Olaf Lenz} 
\institute{Institut für Computerphysik\\Universit\"at Stuttgart}
\date{March 17-21, 2014}

\setbeamertemplate{footline}{}
\begin{frame}
  \titlepage
\end {frame}
\setbeamertemplate{footline}[icp]

\begin{frame}
  \frametitle{Outline}
  \begin{itemize}
  \item Definition of classes
  \item Lifetime of objects
  \item Constructor, destructor
  \item Default and copy constructor
  \item Encapsulation: private, public
  \item Friends
  \item Function and operator overloading
  \item Inheritance
  \item Protected encapsulation
  \item Polymorphism and virtual functions
  \item Abstract base classes and pure virtual functions
  \item Multiple inheritance
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Definition of classes}

  \begin{onlyenv}<1> \tikzumlset{font=\tiny}
    \begin{tikzpicture}
      \umlclass{Player}{
        playerNum : int\\
        name : string }{
        setPlayerNum(playerNum : num)\\
        getPlayerNum() : int\\[\medskipamount]
        doNextMove(piece : Piece, event : Event, out message :
        string) }
    \end{tikzpicture}
    ~~becomes
    \medskip
\end{onlyenv}

\begin{lstlisting}[emph={class,public}]
class Player {
public:
  int playerNum;
  string name;
  // setter and getter for the player number
  void setPlayerNum(int playerNum);
  int getPlayerNum();
  // perform the next move
  void doNextMove(Piece &piece, Event event, string &msg);
};
\end{lstlisting}

\begin{onlyenv}<2>
  \begin{itemize}
  \item \alert{remember semicolon at the end of the definition!}
  \item \lstinline!public! will be explained later
  \item output parameters translate to references (\lstinline!message!)
  \item no separator between member functions and variables required
  \item good style to separate them
  \end{itemize}
\end{onlyenv}
\end{frame}

\begin{frame}[fragile]
  \frametitle{(Forward) declaration}

  \begin{tikzpicture}
    \umlclass[type=abstract]{Player}{
    }{}

    \umlclass[x=6]{Piece}{
    }{}
    \umlunicompo[arg2=*,anchors=10 and 170]{Player}{Piece}
    \umluniassoc[arg2=1,anchors=-170 and -10]{Piece}{Player}
  \end{tikzpicture}
  ~~becomes
  \medskip

\begin{lstlisting}
class Player; // <-
class Piece {
  Player &player;
};
class Player {
  vector< Piece* > pieces;
};
\end{lstlisting}

  \begin{itemize}
  \item classes can be (forward) \alert{declared} (here: \lstinline!Player!)
  \item necessary if \lstinline!Piece! uses \lstinline!Player!
    and vice versa
  \item ok as long as only addresses are needed (pointers or references)
  \item usually, forward declare higher level classes
  \end{itemize}
\end{frame}

\begin{frame}[squeeze,fragile]
  \frametitle{Declaration of member functions}

\begin{lstlisting}[title={\footnotesize\bfseries Player.hpp}]
#ifndef PLAYER_HPP
#define PLAYER_HPP
class Player {
  void doNextMove(Piece &piece, Event ev, string &msg);
  int getPlayerNum() { return playerNum; }
};
#endif
\end{lstlisting}

    \begin{itemize}
    \item member functions are declared inside the class definition
    \item definition of \emph{short} functions in class (\lstinline[inline]!inline!)
    \item class definitions are usually placed in header file `\textit{class}.hpp'
    \item preprocessor guards to avoid double definition of class
    \end{itemize}
\end{frame}

\begin{frame}[squeeze,fragile]
  \frametitle{Definition of member functions}

\begin{lstlisting}[title={\footnotesize\bfseries Player.hpp}]
class Player {
  void doNextMove(Piece &piece, Event ev, string &msg);
  int getPlayerNum() { return playerNum; }
};
\end{lstlisting}

\begin{lstlisting}[title={\footnotesize\bfseries Player.cpp}]
void Player::doNextMove(Piece &piece, Event ev, string &msg) {
  piece.tryStep(ev, msg);
}
\end{lstlisting}

    \begin{itemize}
    \item full member identifier: \lstinline[style=inline,deletekeywords={class}]!class::member!
    \item only one definition per member function (including inline def's)
    \item member function definitions in separate cpp-file `\textit{class}.cpp'
    \end{itemize}
\end{frame}

\begin{frame}[squeeze,fragile]
  \frametitle{Scope of member functions}

\begin{lstlisting}
class Player {
  Piece *currentPiece;
  void doNextMove(Piece &piece, Event ev, string &msg) {
    currentPiece = &piece;
    if (piece.tryStep(ev, msg))
      currentPiece->doStep(ev, msg);
  }
};
\end{lstlisting}

  \begin{itemize}
  \item call member functions by \lstinline[style=inline]!object.function()!
  \item also access to member variables, if public
  \item compare C struct syntax
  \item pointers can be dereferenced using `->' like in C
  \item only almost true due to operator overloading (later)
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Referring to yourself: this}

\begin{lstlisting}[emph={this}]
void Piece::register() {
  board.addPiece(this);
}
void Piece::escape() {
  board.removePiece(this);
}
\end{lstlisting}

  \begin{block}{}
    How can an object refer to ``itself''?\\
    \emph{\lstinline!Piece! registers itself with the board}
  \end{block}

  \begin{block}{Answer}
    In a member function, \lstinline!this! always points to the
    current object
  \end{block}

  \begin{itemize}
  \item \lstinline!this! is a pointer, although guaranteed to be valid
  \item there is no reference to the current object, use \lstinline!*this!
    if necessary
  \end{itemize}
\end{frame}

\begin{frame}[fragile,fragile]
  \frametitle{Creation and life time of objects}

  \begin{onlyenv}<1>
\begin{lstlisting}
int test() {
  for (int i = 0; i < 10; ++i) {
    Piece piece;
    piece.register();
    // here, piece is destroyed automatically
  }
}
\end{lstlisting}

    \begin{itemize}
    \item objects are created like variables in C by giving type and name
    \item ...and die at the end of the scope (code block)
    \item exception: static variables like in C
    \item therefore, after the loop above, not a single piece exists
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<2>
\begin{lstlisting}
int test() {
  for (int i = 0; i < 10; ++i) {
    Piece *piecePtr = new Piece;
    piecePtr->register();
    delete piecePtr;
    // without the older pieces remain, inaccessible
  }
}
\end{lstlisting}

    \begin{itemize}
    \item objects can also be created by \lstinline!new!
    \item these live till they are destroyed by explicitely calling
      \lstinline!delete! \emph{once}
    \item there should be \emph{one} responsible object for
      destroying, the \alert{owner}
    \item more than one owner leads to segmentation faults
    \item no owner (and therefore no \lstinline!delete!) to memory leaks
    \item in OOP, ownership frequently changes, making things
      difficult
    \end{itemize}
  \end{onlyenv}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Constructor}

\begin{lstlisting}
class Game {
  Board* board;
  vector< Player* > players;
public:
  Game(int numPlayers = 2): board(0) {
    for (int i = 0; i < numPlayers; ++i)
      players.push_back(new Player);
  }
};
\end{lstlisting}

  \begin{itemize}
  \item constructor is a function initializing an object
  \item has the same name as the classs (``Class::Class'')
  \item can take arguments that specify what to create
  \item there can be several constructors (see overloading)
  \item initialize member variables after colon (here: \lstinline!board!)
  \item multiple variables separated by comma
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Destructor}

\begin{lstlisting}
class Game {
  Board* board;
  vector< Player* > players;
public:
  ~Game() {
    for (auto player: players) delete player;
    delete board;
  }
};
\end{lstlisting}

  \begin{itemize}
  \item destructor is called right \emph{before} unallocating the
    memory
  \item should clean up
  \item in particular destroy all owned objects
  \item inform other objects that have pointers to this object
  \item the destructor cannot take arguments
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Default and copy constructor}

\begin{lstlisting}
class Game {
public:
  Game()      { /* call default constructors of all objects */ }
  Game(const Game &src)  { /* use copy constructors instead */ }
};
\end{lstlisting}

  \begin{itemize}
  \item two \emph{predefined} constructors:
    \begin{itemize}
    \item Default constructor \lstinline!Class::Class()!
      \begin{itemize}
      \item only defined if no user-defined constructor declared
      \item initializes all objects using their default constructor
      \item plain old data types (char, int, ...) are uninitialized
      \end{itemize}
    \item Copy constructor \lstinline!Class::Class(const Class &src)!
      \begin{itemize}
      \item copies all member variables from object \lstinline!src!
      \item for pointers often a bad idea - duplicates ownership
      \end{itemize}
    \end{itemize}
  \item to avoid automatic generation, declare constructors
    \emph{private}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Static class members}

\begin{lstlisting}[emph={static,getCount}]
class CountOccurance {
  static int cnt;
public:
  CountOccurance() { ++cnt; }
  ~CountOccurance() { --cnt; }
  static int getCount() { return cnt; }
};
int CountOccurance::cnt = 0;
int main() {
  new CountOccurance();
  cout << CountOccurance::getCount() << endl;
  return 0;
}
\end{lstlisting}

  \begin{itemize}
  \item class members (functions and variables) can be static
  \item variable needs to be defined once somewhere (not in header!)
  \item static member functions are called with \emph{class} prefix
  \end{itemize}

\end{frame}

\begin{frame}[fragile]
  \frametitle{Function overloading}
  
  \begin{center}
    \begin{tikzpicture}
      \umlclass[type=abstract]{Tile}{ }{
        \umlvirt{tryStep(piece : Piece, how : Event, out message : string)}\\
        \umlvirt{tryStep(monster : Monster, how : Event, out message : string)}\\
      }
    \end{tikzpicture}
  \end{center}
  becomes
\begin{lstlisting}
class Tile {
public:
  void tryStep(Piece &piece, Event how, string &message);
  void tryStep(Monster &monster, Event how, string &message);
};
\end{lstlisting}

  \begin{itemize}
  \item several methods with same name but different signature
  \item signature is formed by the types of all taken arguments
  \item return value is \emph{not} part of the signature
  \item also global (C-style) functions can be overloaded
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Default arguments}
  
\begin{lstlisting}
class QGameBoard
{
public:
  QGameBoard(int tileSize = 20, QWidget *parent = 0);
  void setTileSize(int sizeX, int sizeY = 0);
};
\end{lstlisting}

  \begin{itemize}
  \item another kind of function overloading are default arguments
  \item values get predefined values if not specified when calling
  \item always starts from the back:\\
    \lstinline!QGameBoard(30)! is ok: \lstinline!parent = 0!\\
    \lstinline!QGameBoard(otherWidget)! not:
    \lstinline!parent = 0!, the pointer \lstinline!otherWidget! is cast into an
    integer for \lstinline!tileSize!
  \item here, this also overloads the default constructor
    \lstinline!QGameBoard()!
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Operator overloading}

  \begin{itemize}
  \item most operators can be overloaded
  \item abuse can lead to \emph{big} confusion
  \item obviously relies heavily on signatures
  \end{itemize}

Some examples:
\begin{lstlisting}
// output using iostream
ostream &operator<<(ostream &stream, const Class &a);
// assignment of any type:
Class &Class::operator=(OtherClass b);
// sum (product, difference,... analogously):
RetClass Class::operator+(OtherClass b);
// ++object:
Class &Class::operator++();
// object++ (yes, they can differ):
Class Class::operator++(int); // int is a dummy argument!
// for objects representing functions:
RetClass Class::operator()(ParamClass a, ParamClass b, ...);
\end{lstlisting}

\end{frame}

\begin{frame}[fragile]
  \frametitle{Inheritance}
  \tikzumlset{font=\tiny}
  \begin{tikzpicture}
    \umlclass{Tile}{
      x, y : int
    }{
      getPosition(out int, out int)\\
      tryStep(...)
    }
      
    \umlclass[x=5]{WallTile}{
    }{
      tryStep(...)
    }

    \umlinherit{WallTile}{Tile}
  \end{tikzpicture}
  ~~becomes
  \medskip
\begin{lstlisting}[emph={public}]
class Tile {
  int x, y;
  void getPosition(int x, int y);
  virtual void tryStep(Piece &piece, Event how, string &message);
};
class WallTile : public Tile {
  virtual void tryStep(Piece &piece, Event how, string &message);
};
\end{lstlisting}

  \begin{itemize}
  \item derived classes declare super classes after class name and colon
  \item getPosition also exists in WallTile (\alert{inherited} from \lstinline!Tile!)
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Constructor and base classes}

\begin{lstlisting}
class Tile {
public:
  Tile();
  ~Tile();
};
class WallTile: public Tile {
public:
  WallTile(): Tile() { }  
  WallTile(int) { }  
  ~WallTile();
};
\end{lstlisting}

  \begin{onlyenv}<1>
    \begin{itemize}
    \item base classes are initialized in constructor following colon\\
      (like member variables)
    \item using default constructors can be omitted (\lstinline!WallTile(int)!)
    \item arguments to constructors can be computed freely
    \item base classes are automatically destroyed
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<2>
    \begin{itemize}
    \item order follows matryoshka principle: from general to special
    \item base class constructors come \emph{first}
    \item base class destructors come \emph{last}
    \end{itemize}
  \end{onlyenv}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Encapsulation: private, public, protected}

  \begin{onlyenv}<1>
    \tikzumlset{font=\tiny}
    \begin{center}
      \begin{tikzpicture}
        \umlclass[type=abstract]{Tile}{
          - notvisible: bool\\
          \# x, y: int }{ + getPosition(out int, out int) }
        \umlemptyclass[y=-1,x=3.5]{WallTile}
        \umlemptyclass[x=6]{Player}
      
        \umlinherit{WallTile}{Tile} \umldep{Player}{WallTile}
        \umldep{Player}{Tile}
      \end{tikzpicture}
    \end{center}
    \vspace{-\baselineskip}
    becomes
    \medskip
  \end{onlyenv}

\begin{lstlisting}[emph={public,private,protected}]
class Tile {
private:
  bool notvisible;
protected:
  int x, y;
public:
  void getPosition(int x, int y);
};
class WallTile : public Tile { };
\end{lstlisting}

  \begin{onlyenv}<2>
    \begin{itemize}
    \item \alert{private} members are only accessible to members of
      the class
    \item \alert{protected} is also accessible to members of derived
      classes
    \item \alert{public} members are visible to all
    \item no difference between function and variable members
    \item a non-public destructor hinders destroying objects ---
      rarely a good idea
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<3>
    \begin{itemize}
    \item \lstinline!notvisible! is only accessible to
      \lstinline!Tile!  instances (\alert{private})
    \item \lstinline!x! and !y! are accessible to instances of classes
      \lstinline!Tile! and \lstinline!WallTile! (\alert{protected})
    \item \lstinline!getPosition! is visible to all, including
      e.g. \lstinline!Player! instances (\alert{public})
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<4>
    \begin{itemize}
    \item base classes also have a visibility (private, protected or public)
    \item most stringent visibility applies
    \item private members never become protected or public
    \item protected members never become public
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<5>
    \begin{itemize}
    \item objects are always also base class objects\\
      \emph{(a \lstinline!WallTile! is a \lstinline!Tile!, a
      \lstinline!Mammal! an \lstinline!Animal!)}\\
      $\implies$ \alert{base class (interface) should be public}
    \item ... if not strong reasons against
    \item Qt: $\approx 2300$ public inheritances, $\approx 30$ private
      and $\approx 10$ protected
    \end{itemize}
  \end{onlyenv}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Friends}

\begin{lstlisting}[emph={friend}]
class Hidden {
private:
  Hidden();
  Hidden(const Hidden &);
  ~Hidden();
  friend ostream &operator<<(ostream &, const Hidden &);
  friend class HiddenFactory;
};
\end{lstlisting}
  \begin{itemize}
  \item \lstinline[style=inline]!friend! overrides any visibility restrictions
  \item friends are declared in the class definition\\
    $\implies$ you cannot inject friends
  \item friends can be classes or (member) functions
  \item common use is allowing the \lstinline[style=inline]!<<!-operator to access inner parts
  \item here, it enforces to use \lstinline[style=inline]!HiddenFactory! to create objects
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Polymorphism: virtual functions}

\begin{lstlisting}[emph={virtual}]
class Tile {
  virtual string whatAmI() { return "tile"; }
};
class FloorTile: public Tile {
  virtual string whatAmI() { return "floor tile"; }
};
void informMe(Tile &tile) {
  cout << "this is a " << tile.whatAmI() << endl;
}
\end{lstlisting}

\begin{itemize}
\item without virtual, \lstinline[style=inline]!informMe! always says ``this is a tile''
\item \lstinline!virtual! declares a function to be \alert{polymorph}
\item acts differently depending on the \emph{actual} type of the object
\item \lstinline!Tile::whatAmI! is different for a
	\lstinline!Tile! and a \lstinline!WallTile!
\item forgetting to define a virtual function leads to undefined
      references to the `vtable for class'
\end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Pure virtual functions and abstract base classes}

\begin{lstlisting}
class Tile {
  virtual string whatAmI() = 0;
};
class FloorTile: public Tile {
  virtual string whatAmI() { return "floor tile"; }
};
int main() {
  // Error! Abstract class, cannot be instantiated
  Tile tile;
}
\end{lstlisting}

    \begin{itemize}
    \item purely virtual (abstract) functions are only defined for derived classes
    \item classes with purely virtual functions are \alert{abstract}
      classes
    \item you cannot create objects of abstract classes
    \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Multiple inheritance}
  
  \begin{onlyenv}<1>
    \tikzumlset{font=\tiny}
    \begin{tikzpicture}
      \umlclass{Person}{ name: string }{} \umlclass[x=5]{Horse}{
        tailLength : float }{} \umlemptyclass[x=2.5,y=-1]{Centaur}

      \umlinherit{Centaur}{Person} \umlinherit{Centaur}{Horse}
    \end{tikzpicture}
    ~~becomes
  \end{onlyenv}

\begin{lstlisting}[emph={public}]
class Person {
  string name;
  Person(const string &n): name(n) {}
};
class Horse {
  float tailLength;
  Horse(float t): tailLength(t) {}
};
class Centaur: public Person, public Horse {
  Centaur(const string &n, float t) : Person(n), Horse(t) {}
};
\end{lstlisting}

  \begin{onlyenv}<2>
    \begin{itemize}
    \item multiple base classes are separated by commas
    \item each carries a visibility (public, protected or private)
    \item class has member variables and functions of both base
      classes
    \end{itemize}
  \end{onlyenv}
\end{frame}

\begin{frame}
  \frametitle{The diamond problem}

  \begin{center}
    \tikzumlset{font=\tiny}
    \begin{tikzpicture}
      \umlclass[x=3]{Person}{
        name: string
      }{
        getNameAndTitle() : string
      }
      \umlclass[y=-2.5]{Student}{}{
        \umlvirt{getNameAndTitle() : string}
      }
      \umlclass[y=-2.5,x=6]{Teacher}{}{
        \umlvirt{getNameAndTitle() : string}
      }
      \umlclass[y=-5,x=3]{StudyingTeacher}{}{
        \umlvirt{getNameAndTitle() : string}
      }
      
      \umlinherit{StudyingTeacher}{Teacher}
      \umlinherit{StudyingTeacher}{Student}
      \umlinherit{Teacher}{Person}
      \umlinherit{Student}{Person}
    \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}[fragile]
  \frametitle{The diamond problem --- solution in C++}

\begin{lstlisting}[basicstyle={\ttfamily\tiny},emph={virtual}]
class Person {
protected: string name;
public:    Person(const string &_name): name(_name) { }
           virtual string getNameAndTitle();
};
class Teacher: virtual public Person {
public:    Teacher(const string &_name): Person(_name) { }
           virtual string getNameAndTitle();
};
class Student: virtual public Person {
public:    Student(const string &_name): Person(_name) { }
           virtual string getNameAndTitle();
};
class StudyingTeacher
  : virtual public Teacher, virtual public Student {
public:    StudyingTeacher(const string &n)
             : Person(n), Teacher(n), Student(n) {}
           virtual string getNameAndTitle();
};
\end{lstlisting}

  \begin{itemize}
  \item declare base classes virtual
  \item requires also to initialize the common ancestor in the constructor
  \item initializiation of ancestor in intermediate classes is skipped
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{constant objects --- const}
  \begin{onlyenv}<1>
\begin{lstlisting}[emph={const}]
const int theNumber = 42;
const char *nonConstantPtrToConstChar = "Hase";
nonConstantPtrToConstChar = "Igel";
const char * const constantPtrToConstChar = "Hase";
const QPointF constantPoint(2, 3);
\end{lstlisting}

    \begin{itemize}
    \item \lstinline!const! objects cannot be changed
    \item can be of any type (plain old data type, pointer, struct or class)
    \item the compiler generates code based on this assumption!
    \item be careful with pointers
      \begin{itemize}
      \item \lstinline!const! \emph{on the left} binds to the data
        type pointed to ---\\
        you cannot change the data, but point to different data
      \item \lstinline!const! \emph{on the right} makes the address
        pointed to unchangeable
      \end{itemize}
    \item C-string constants are of type \lstinline!const char!,
      i.e. cannot be changed after compilation
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<2>
\begin{lstlisting}[emph={const}]
class Class {
public:
  const int myInitialValue;
  Class(int n): myInitialValue(n) { }
};
\end{lstlisting}

    \begin{itemize}
    \item classes can define constant member variables
    \item even if public, they cannot be changed by any object
    \item they have to be initialized in the initializer list of the
      constructor(s) (after the colon)
    \end{itemize}

  \end{onlyenv}

  \begin{onlyenv}<3>
\begin{lstlisting}[emph={const}]
class Class {
  int myNumber;
public:
  Class(int n) { myNumber = n; }
  void setMyNumber(int i) { myNumber = i; }
  int getMyNumber()  const { return myNumber; }
};
\end{lstlisting}
    
    \begin{itemize}
    \item classes need to be \alert{const-aware} to make use of constant objects
    \item on constant objects, only member functions declared
      \lstinline!const! can be called\\
      \emph{here, only \lstinline!getMyNumber!, but not \lstinline!setMyNumber!}
    \item \lstinline!const! functions can also called on non-const objects
    \item \lstinline!const! member functions cannot change
      any member variables
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<4>
\begin{lstlisting}[emph={const}]
class Class {
public:
  void test()       { cout << "nonconst" << endl; }
  void test() const { cout << "const" << endl; }
};
Class changeable(1);
const Class unchangeable(2);
\end{lstlisting}

    \begin{itemize}
    \item \lstinline!const! is part of the \emph{signature}
    \item \lstinline!const! and non-\lstinline!const! versions of the
      same function can exist
    \item here, \lstinline!changeable.test()! will output
      ``nonconst'', \lstinline!unchangeable.test()! will output
      ``const''
    \item return values can differ (used frequently in STL)
\begin{lstlisting}
class vector {
  iterator       begin();
  const_iterator begin() const;
\end{lstlisting}
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<5>
    \begin{block}{Pros}
      \vspace{-\bigskipamount}
      \begin{itemize}
      \item const-aware programs are less error-prone
      \item functions that should not change anything cannot do so
      \item allows the compiler to optimize more aggressively
      \item if writing libraries, const-awareness is necessary for
        others to be able to program const-aware
      \end{itemize}
    \end{block}

    \begin{block}{Cons}
      \vspace{-\bigskipamount}
      \begin{itemize}
      \item hard to interface non-const-aware code (e.g. plain C)
      \item leads often to code duplication
      \item vector e.g. needs const and non-const iterators, which
        essentially do the same
      \item templates reduce programming overhead, but not
        code size
      \end{itemize}
    \end{block}
  \end{onlyenv}
\end{frame}

\begin{frame}[fragile]
  \frametitle{\lstinline!dynamic_cast! and \lstinline!static_cast!}

  \begin{onlyenv}<1>
\begin{lstlisting}[emph={static_cast}]
class Tile {

};
class WallTile: public Tile { };

int main() {
  Tile *tile = new Tile;
  WallTile *wall = new WallTile;
  Tile *tilePtr = static_cast<Tile *>(wall);
  WallTile *wallPtr = static_cast<WallTile *>(tile);
}
\end{lstlisting}
    \begin{itemize}
    \item \lstinline!static_cast! on class pointers only checks
      compatibility (target class should be super or derived class)
    \item ends badly if a base object gets cast to a derived object
    \item here, \lstinline!wallPtr! should not be used!
    \item still better than C-like casting, which \alert{you shall not use!}
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<2>
\begin{lstlisting}[emph={dynamic_cast}]
class Tile {

};
class WallTile: public Tile { };

int main() {
  Tile *tile = new Tile;
  WallTile *wall = new WallTile;
  Tile *tilePtr = dynamic_cast<Tile *>(wall);
  WallTile *wallPtr = dynamic_cast<WallTile *>(tile);
}
\end{lstlisting}
    \begin{itemize}
    \item \lstinline!dynamic_cast! checks full compatibility
    \item here, the cast to \lstinline!WallTile! for
      \lstinline!wallPtr! causes a compiler error
    \item without polymorphism only casting to base classes possible
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<3>
\begin{lstlisting}[emph={dynamic_cast,makeMePolymorph}]
class Tile {
  virtual void makeMePolymorph() {};
};
class WallTile: public Tile { };

int main() {
  Tile *tile = new Tile;
  WallTile *wall = new WallTile;
  Tile *tilePtr = dynamic_cast<Tile *>(wall);
  WallTile *wallPtr = dynamic_cast<WallTile *>(tile);
  if (wallPtr) use(wallPtr);
}
\end{lstlisting}
    \begin{itemize}
    \item on polymorphic classes (with at least one virtual function)
      \lstinline!dynamic_cast! returns 0 if the cast is impossible
    \item here, the cast to \lstinline!WallTile! for
      \lstinline!wallPtr! is syntactically ok
    \item but \lstinline!wallPtr! will be 0
    \end{itemize}
  \end{onlyenv}

\end{frame}

\begin{frame}[fragile]
  \frametitle{Example of using \lstinline!dynamic_cast!}
\begin{lstlisting}
for (auto it : content->items()) {
  QGraphicsTextItem *item =
    dynamic_cast<QGraphicsTextItem *>(it);
  if (item)
    item->setFont(QFont("Helvetica"));
}
\end{lstlisting}

  \begin{itemize}
  \item working with lists of different items (e.g. \lstinline!QGraphicsScene!)
  \item all elements are pointers to a common base class (here:
    \lstinline!QGraphicsItem!)
  \item for each item, you can check whether it is e.g. a
    \lstinline!QGraphicsTextItem! and then change the font
  \end{itemize}  
\end{frame}

\begin{frame}[fragile]
  \frametitle{const -- non-const conversion: \lstinline!const_cast!}

\begin{lstlisting}
const char *unchangeable;
char *changeable =
          const_cast<char *>(unchangeable);

char *changeable2;
const char *unchangeable2 =
          const_cast<const char *>(changeable2);
\end{lstlisting}

  \begin{itemize}
  \item \lstinline!const_cast! converts between const and non-const variables
  \item if \lstinline!unchangeable! points to a true C-string, writing
    to \lstinline!changeable! will still cause a segmentation fault
  \item only use: interfacing with C-libraries that are not const-aware
  \item you need to know that the function doesn't change anything
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Automatic type conversion}

  \begin{onlyenv}<1>
\begin{lstlisting}
class String {
  string value;
public:
  String(const string &val): value(val) {}
};
int main()
{
  String s("Hallo");
}
\end{lstlisting}

    \begin{itemize}
    \item this code works due to \alert{automatic type conversion}
    \item compiler sees that String's constructor takes a
      \lstinline!string! instance
    \item it tries to convert the \lstinline!const char *! to a string
      using any string's constructors
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<2>
\begin{lstlisting}
class MyString {
public:
  MyString(const char *val);
};
class User {
public:
  User(MyString bla);
  User(string bla);
};
int main() {
  User user("Hallo");
}
\end{lstlisting}

    \begin{itemize}
    \item creating \lstinline!user! fails since the conversion is
      \alert{ambigous}
    \item the intermediate class could equally be \lstinline!MyString!
      or \lstinline!string!
    \item both are accepted by the User class
    \item both can be constructed from \lstinline!const char *!
    \end{itemize}
  \end{onlyenv}

  \begin{onlyenv}<3>
\begin{lstlisting}[emph={explicit}]
class MyString {
public:
  explicit MyString(const char *val);
};
class User {
public:
  User(MyString bla);
  User(string bla);
};
int main() {
  User user("Hallo");
  User user2(MyString("Welt"));
}
\end{lstlisting}

    \begin{itemize}
    \item \lstinline!explicit! excludes constructors from
      consideration for implicit conversion
    \item you can use them by explicitely (!) calling the constructor
    \item solves the ambiguity
    \end{itemize}
  \end{onlyenv}
\end{frame}

\begin{frame}[fragile]
  \frametitle{The \lstinline!auto! data type}

\begin{lstlisting}
void print(const vector<int> &v) {
  for (auto value: v) 
    cout << value << endl;
}
const auto theNumber = 5;
\end{lstlisting}

  \begin{itemize}
  \item C++11 adds the data type \lstinline!auto!
  \item determines the data type from the right hand sides datatype
  \item works if the right hand side has a unique type
    \begin{itemize}
    \item constants and other variables
    \item return values of functions (return value cannot be overloaded!)
    \end{itemize}
  \item in particular useful with range-based for (see STL, Olaf)
  \end{itemize}
\end{frame}
\end{document}
